---
title: "HW4"
author: "Nathan Bick"
date: '2022-10-04'
output: pdf_document
---

```{r setup}
options(scipen=999)
library(dplyr)
library(stringr)
```

# Problem 1

Download the datafile `economic_data.txt`. See the datafile for details regarding the column values and meaning, note that the datafile has six covariates $(A_1,A_2,...,A_6)$. Consider the multiple linear model, and find the best fit for this model. 

## Response

Economic and unemployment data were recorded.

There are 16 rows of data.  The data include:

* I,  the index;
* A1, the percentage price deflation;
* A2, the GNP in millions of dollars;
* A3, the number of unemployed;
* A4, the number of people employed by the military;
* A5, the number of people over 14;
* A6, the year
* Y,  the number of people employed.

A3-A5 are in units of thousands

We seek a model of the form:

$$Y = \beta_0 + A_1 \beta_1 + A_2 \beta_2 + A_3 \beta_3 + A_4 \beta_4 + A_5 \beta_5 + A_6 \beta_6$$.

First we read in the data and do some data cleaning. This is followed by two portions

```{r}
# read in the data
economic_data <- read.csv(
  "~/Documents/Graduate School/MATH504 - Numerical Methods/Homework/HW4/economic_data.txt",
  sep = "\t",
  skip=35,
  header = FALSE)

# we see that there are some issues with the white spaces, so we do data cleaning. 
# this outputs a dataframe which can be used for the R native linear model. 
economic_data_clean <- economic_data %>%
  mutate(V1 = gsub("[[:space:]]+", ",", trimws(V1))) %>% 
  tidyr::separate(
    sep = ",",
    col =  V1,
    into =  c("I","A1","A2","A3","A4","A5","A6","Y")
  ) %>% mutate_all(as.numeric)

```

we need matrices for our manual solution, so we get those here.

```{r} 
X = economic_data_clean %>% select(-I,-Y) %>% as.matrix()
A = economic_data_clean %>% 
  mutate(one = 1)%>% select(one,A1,A2,A3,A4,A5,A6,-I,-Y) %>% as.matrix()
y = economic_data_clean %>% select(Y) %>% as.matrix()
```


We have two ways to solve this. First, we can use what we learned in the lecture to directly solve via the following minimization problem:

$$min || A\beta - y||^2$$ over $\beta$. This corresponds to the following formula.

$$\beta^*=(A^TA)^{-1}A^Ty$$

We solve this formula with the following R code. This gives us the set of coefficients $\beta$ vector.
 
```{r}
solve((t(A) %*% A), tol = 1e-21) %*% t(A) %*% y

```

We can check our answer using the internal `lm()` function that does what we just did by hand. This gives us the same $\beta$ vector. We see the results are the same.

```{r}
lm(
  data = economic_data_clean,
  formula = Y ~ A1 + A2 + A3 + A4 + A5 + A6)
```

We now see that these coefficients are the same in both methods. 
